---
title: "Understanding Youth Crime Patterns in Victoria: A Decade of Change (2015–2024)"
author: "- Anushka Sauran"
output:
  flexdashboard::flex_dashboard:
    storyboard: true
    orientation: columns
    vertical_layout: fill
    theme: united
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = "center", warning = FALSE, message = FALSE)

# Loading the necessary libraries for data analysis and visualization
library(tidyverse)
library(flexdashboard)
library(gt)
library(ggplot2)
library(readxl)
library(plotly)
library(DT)
library(scales)
library(viridis)
```

```{r}
#Data Preparation and Processing
# This section transforms raw crime statistics into meaningful insights about youth offending patterns.

# Importing the youth crime excel file
crime_data_file <- "youth crime 2015-2024.xlsx"

# Extract youth offense trends organized by demographics
# This sheet contains age-based breakdowns that help us understand developmental
youth_demographics_raw <- read_excel(crime_data_file, sheet = "Table 02")

# Focus on the most vulnerable age groups where intervention can be most effective
# These age ranges represent critical developmental periods
vulnerable_age_groups <- youth_demographics_raw %>%
  filter(`Age Group` %in% c("10-14 years", "15-17 years", "18-19 years")) %>%
  select(Year, `Age Group`, `Alleged Offender Incidents`) %>%
  group_by(Year, `Age Group`) %>%
  # Aggregate incidents to get clear yearly totals for each age group
  summarise(yearly_incidents = sum(`Alleged Offender Incidents`, na.rm = TRUE)) %>%
  ungroup()  

# Import offense type data to understand what crimes young people are committing
# This helps inform prevention strategies
offense_patterns_raw <- read_excel(crime_data_file, sheet = "Table 04")

# Analyze the most recent crime patterns for current policy relevance
# Focusing on 15-17 year olds as they show highest incident rates
current_offense_breakdown <- offense_patterns_raw %>%
  filter(Year == 2024, `Age Group` == "15-17 years") %>%
  group_by(`Offence Division`) %>%
  # Calculate both raw numbers and percentages for comprehensive understanding
  summarise(incident_count = sum(`Alleged Offender Incidents`, na.rm = TRUE)) %>%
  mutate(percentage_of_total = incident_count / sum(incident_count) * 100)

# Import detailed age-specific data for granular analysis
# This provides insights into how offending patterns change year by year
individual_age_data <- read_excel(crime_data_file, sheet = "Table 07")

# Clean and standardize the individual age data for consistent analysis
detailed_age_patterns <- individual_age_data %>%
  # Focus on individual ages within our target range
  filter(`Age Group` %in% as.character(10:19),
         # Use specific gender categories to avoid double-counting totals
         Sex %in% c("Males", "Females")) %>%  
  rename(individual_age = `Age Group`) %>%
  # Standardize gender labels for cleaner visualization
  mutate(gender_category = case_when(
    Sex == "Males" ~ "Male",
    Sex == "Females" ~ "Female",
    TRUE ~ Sex
  )) %>%
  group_by(individual_age, gender_category) %>%
  # Sum incidents across all offense types for each age-gender combination
  summarise(total_incidents = sum(`Alleged Offender Incidents`, na.rm = TRUE)) %>%
  ungroup()

# Analyze how different types of crimes have evolved over time
# This helps identify emerging trends and persistent problems
crime_evolution_patterns <- offense_patterns_raw %>%
  filter(`Age Group` %in% c("10-14 years", "15-17 years", "18-19 years")) %>%
  group_by(Year, `Offence Division`) %>%
  summarise(annual_incidents = sum(`Alleged Offender Incidents`, na.rm = TRUE)) %>%
  ungroup() %>%
  # Filter to focus on crime types with substantial impact
  group_by(`Offence Division`) %>%
  filter(sum(annual_incidents) > 1000) %>%
  ungroup()

# Examine gender differences across age groups for targeted interventions
# Understanding these patterns helps design appropriate prevention programs
gender_age_distribution <- youth_demographics_raw %>%
  filter(`Age Group` %in% c("10-14 years", "15-17 years", "18-19 years"),
         Year == 2024) %>%
  group_by(`Age Group`, Sex) %>%
  summarise(total_annual_incidents = sum(`Alleged Offender Incidents`, na.rm = TRUE)) %>%
  ungroup()

# Calculate overall trends to understand the bigger picture context
# This provides the foundation for understanding whether youth crime is increasing or decreasing
overall_youth_trends <- vulnerable_age_groups %>%
  group_by(Year) %>%
  summarise(combined_youth_incidents = sum(yearly_incidents)) %>%
  ungroup() %>%
  # Calculate year-over-year changes to identify trend direction and magnitude
  mutate(annual_change = combined_youth_incidents - lag(combined_youth_incidents),
         percentage_change = (annual_change / lag(combined_youth_incidents)) * 100)

# Create focused gender analysis for the most recent data
# This helps understand current gender disparities in youth offending
recent_gender_patterns <- youth_demographics_raw %>%
  filter(Year == 2024, 
         `Age Group` %in% c("10-14 years", "15-17 years", "18-19 years"),
         # Ensure we capture all gender-specific data accurately
         Sex %in% c("Males", "Females")) %>%  
  group_by(Sex) %>%
  summarise(gender_incident_total = sum(`Alleged Offender Incidents`, na.rm = TRUE)) %>%
  ungroup() %>%
  # Create cleaner labels for presentation
  mutate(gender_label = case_when(
    Sex == "Males" ~ "Male",
    Sex == "Females" ~ "Female",
    TRUE ~ Sex
  ))

```


Slide 1 {data-orientation=columns}
=====================================

**Opening Story: Why Youth Crime Matters to All of Us**

Youth crime affects every Victorian community, from suburban families to urban neighborhoods. When young people get involved in criminal activity, it represents both a failure of support systems and an opportunity for positive intervention. This analysis examines a decade of data about young offenders aged 10-19, seeking to understand patterns that can guide more effective prevention and intervention strategies.

**The Questions We're Investigating:**
- How have youth offending patterns shifted over this transformative decade?
- Which young people are most at risk, and when do they need help most?
- What specific behaviors are driving these statistics?
- How do gender and age intersect to create different risk profiles?

```{r}

# Calculate key figures that tell the immediate story
incidents_2024 <- sum(vulnerable_age_groups$yearly_incidents[vulnerable_age_groups$Year == 2024])
incidents_2015 <- sum(vulnerable_age_groups$yearly_incidents[vulnerable_age_groups$Year == 2015])
decade_change <- incidents_2024 - incidents_2015

# Create an accessible summary table for initial context
story_opening_stats <- data.frame(
  KeyStatistic = c("Young People Involved in Crime (2024)", 
                   "Young People Involved in Crime (2015)", 
                   "Net Change Over Decade"),
  RawNumbers = c(incidents_2024, incidents_2015, decade_change)
)

# Format the table for clear presentation
story_opening_stats %>%
  gt() %>%
  fmt_number(columns = RawNumbers, use_seps = TRUE) %>%
  tab_header(title = md("**The Youth Crime Situation at a Glance**")) %>%
  tab_options(
    table.font.size = "medium",
    heading.align = "left",
    row.striping.background_color = "#F9F9F9"
  )

```
**Where This Data Comes From:** Crime Statistics Agency Victoria. (2024). Alleged offender incidents by age and sex, Victoria, 2015–2024 [Data set]. Victorian Government. https://www.crimestatistics.vic.gov.au/crime-statistics/latest-victorian-crime-data/download-data

**Important Note:** Each "incident" represents a young person's interaction with law enforcement, not necessarily a conviction. This data helps us understand where intervention might be most needed.

Slide 2 {data-orientation=columns}
=====================================

Row {data-height=200}
-------------------------------------

**The Decade Overview: Understanding the Bigger Picture (2015-2024)**

**What This Timeline Reveals:**

- Youth crime reached an unprecedented peak in 2020 with over 61,000 incidents, coinciding with early pandemic disruptions
- 2018 actually showed relatively low incidents (~51,000), suggesting successful interventions during that period
- 2021 saw the decade's lowest point (~45,500) as pandemic restrictions fully took effect
- 2024 shows the highest incidents in the entire decade at 65,000+, representing a concerning escalation
- The pattern shows dramatic swings rather than gradual trends, indicating youth crime responds rapidly to social changes

**Why These Patterns Matter:** The volatile nature of these trends suggests that youth crime is highly sensitive to social, economic, and policy changes. The 2024 peak represents an urgent crisis requiring immediate attention.

Row {data-height=500}
-------------------------------------

```{r}

# Create an engaging visualization that shows the human story behind the numbers
p1 <- ggplot(overall_youth_trends, aes(x = factor(Year), y = combined_youth_incidents)) +
  
  # Use a line to show continuity over time
  geom_line(group = 1, color = "#7F55B1", size = 2, alpha = 0.5) +
  
  # Points emphasize individual years and make data more readable
  geom_point(color = "#AF3E3E", size = 3) +
  scale_y_continuous(labels = comma_format()) +
  labs(title = "A Decade of Youth Crime: The Full Story (2015–2024)",
       subtitle = "All youth aged 10–19 years combined",
       x = "Year", y = "Young People Involved in Criminal Incidents") +
  theme_minimal() +
  theme(plot.title = element_text(size = 12, family = "Arial Black"),
        panel.grid.minor = element_blank())

ggplotly(p1) %>%
  layout(title = list(text = paste0("A Decade of Youth Crime: The Full Story (2015–2024)" ,
                                   "<br><sup>All youth aged (10–19 years) combined</sup>")))

```
Slide 3 {data-orientation=columns}
========================================

Row {data-height=200}
-------------------------------------

**Age Groups - Who's Most at Risk?**

**The Developmental Story These Numbers Tell:** 

• **Ages 10-14 (Early Adolescence):** While showing the lowest numbers, these incidents are particularly concerning because they often predict future problems. Early intervention here can prevent escalation.

• **Ages 15-17 (Peak Adolescence):** This group consistently shows the highest incident rates. Brain development research tells us this is when risk-taking peaks while impulse control is still developing.

• **Ages 18-19 (Late Adolescence):** Numbers remain significant but more stable. Young adults in this range are transitioning to independence but may still benefit from youth-focused interventions.

**The Prevention Opportunity:** Understanding these age-based patterns helps communities target resources where they'll have the greatest impact on young lives.

Row {data-height=500}
-------------------------------------

```{r}

# Create a visualization that highlights the human development aspect
p2 <- ggplot(vulnerable_age_groups, aes(x = factor(Year), y = yearly_incidents, color = `Age Group`)) +
  geom_line(size = 1, aes(group = `Age Group`)) +
  geom_point(size = 3) +
  
  # Use colors that reflect the progression from childhood to early adulthood
  scale_color_manual(values = c("10-14 years" = "#789DBC", 
                                "15-17 years" = "#A5B68D", 
                                "18-19 years" = "#FF8A8A")) +
  scale_y_continuous(labels = comma_format()) +
  labs(title = "Youth Crime by Age Group (2015–2024)",
       subtitle = "Different ages face different risks and require different approaches",
       x = "Year", y = "Annual Criminal Incidents",
       color = "Age Group") +
  theme_minimal() +
  theme(plot.title = element_text(size = 12, family = "Arial Black"),
        plot.subtitle = element_text(size = 11),
        legend.position = "bottom")

ggplotly(p2)%>%
  layout(title = list(text = paste0("Youth Crime by Age Group (2015–2024)" ,
                                   "<br><sup>Different ages face different risks and require different approaches</sup>")))

```
Slide 4 {data-orientation=columns}
========================================

Row {data-height=200}
-------------------------------------

**Peak Offending Years - The Teen Crisis**

**Two Different Types of Crisis:** 

**2020 - The Pandemic Surge:** Counter-intuitively, 2020 saw a massive spike in youth crime incidents despite lockdowns. This suggests that pandemic stresses - family tension, economic hardship, disrupted routines - may have actually increased youth criminal behavior initially.

**2024 - The Current Escalation:** This represents the highest youth crime levels in the entire decade, with all age groups contributing to what appears to be a systemic breakdown in youth support systems.

**Understanding the Difference:** While 2020's surge might have been driven by acute pandemic stress, 2024's crisis appears to reflect deeper, more persistent problems in how we're supporting young people in the post-pandemic world.

Row {data-height=500}
-------------------------------------

```{r}

# Analyze the two major peak periods: 2020 and 2024
crisis_comparison <- vulnerable_age_groups %>%
  filter(Year %in% c(2020, 2024)) %>%
  group_by(Year, `Age Group`) %>%
  summarise(crisis_incidents = sum(yearly_incidents)) %>%
  ungroup()

# Visualize both crisis periods to understand their different characteristics
p3 <- ggplot(crisis_comparison, aes(x = `Age Group`, y = crisis_incidents, fill = factor(Year))) +
  geom_col(position = "dodge", alpha = 0.85, width = 0.7) +
  # Use colors that distinguish between the two crisis periods
  scale_fill_manual(values = c("2020" = "#95D2B3", "2024" = "#E1ACAC"),
                    name = "Crisis Year") +
  scale_y_continuous(labels = comma_format()) +
  # Add value labels to emphasize the scale difference
  geom_text(aes(label = comma(round(crisis_incidents))), 
            position = position_dodge(width = 0.7),
            vjust = -0.3, size = 3.5, fontface = "bold") +
  labs(title = "Two Different Crises: 2020 vs 2024 Youth Crime Peaks",
       subtitle = "Comparing the pandemic surge with the current escalation",
       x = "Age Group", 
       y = "Annual Incidents") +
  theme_minimal() +
  theme(plot.title = element_text(size = 12, family = "Arial Black"),
        plot.subtitle = element_text(size = 11),
        legend.position = "bottom")

ggplotly(p3)%>%
  layout(title = list(text = paste0("Two Different Crises: 2020 vs 2024 Youth Crime Peaks" ,
                                   "<br><sup>Comparing the pandemic surge with the current escalation</sup>")))

```
Slide 5 {data-orientation=columns}
========================================

Row {data-height=200}
-------------------------------------

**The Crimes Behind the Statistics: What Are Young People Actually Doing?**

**The Human Story Behind These Categories:**

Rather than just seeing crime statistics, these categories represent different types of struggles and poor decision-making that young people face:

- **Property Crimes:** Often driven by peer pressure, perceived needs, or thrill-seeking behavior
- **Public Order Offenses:** May reflect social conflicts, substance use, or lack of positive outlets
- **Violent Crimes:** While less common, these often stem from unresolved conflicts or learned behaviors

**Implications for Support:** Understanding what young people are actually doing wrong helps communities design more targeted prevention programs that address root causes rather than just consequences.

Row {data-height=300}
-------------------------------------

```{r}

# Create a visualization that shows the types of behaviors driving the statistics
p4 <- plot_ly(current_offense_breakdown, 
              labels = ~`Offence Division`, 
              values = ~incident_count,
              type = 'pie',
              hole = 0.4,  
              textinfo = 'percent',  
              textposition = 'inside',  
              textfont = list(size = 12, color = 'white'),
              
              # Use a color palette that's both accessible and visually appealing
              marker = list(colors = c('#789DBC', '#BACD92', '#E5989B', '#AD88C6', 
                                     '#FFB996', '#819A91'),
                            line = list(color = '#FFFFFF', width = 2))) %>%
  layout(title = list(text = "Understanding Youth Behavior: Crime Types Among 15-17 Year Olds (2024)",
                     font = list(size = 12, family = "Arial Black", color = "black")),
         showlegend = TRUE,
         legend = list(orientation = "v", 
                      x = 1.01, 
                      y = 0.5,
                      font = list(size = 11),
                      itemwidth = 30),
         margin = list(l = 40, r = 220, t = 60, b = 40),
         width = 700,
         height = 500)
p4

```

Slide 6 {data-orientation=columns}
========================================

Row {data-height=200}
-------------------------------------

**Gender Patterns: Why Young Men and Women Follow Different Paths**

**Understanding the Gender Gap:** The significant difference in incident rates between young men and women reflects complex social, developmental, and cultural factors:

• **Social Expectations:** Young men may face different pressures around risk-taking and aggression

• **Peer Influences:** Male peer groups may normalize certain types of risky behavior

• **Expression of Problems:** Young men and women may express underlying issues (trauma, family problems, mental health struggles) in different ways

**Why This Matters for Prevention:** Effective prevention programs need to account for these different pathways into criminal behavior, offering gender-responsive approaches that address the specific risks each group faces.

Row {data-height=500}
-------------------------------------

```{r}

# Ensure proper ordering for clear comparison
recent_gender_patterns$gender_label <- factor(recent_gender_patterns$gender_label, levels = c("Female", "Male"))

# Create a visualization that highlights gender differences thoughtfully
p5 <- ggplot(recent_gender_patterns, aes(x = gender_label, y = gender_incident_total, fill = gender_label)) +
  geom_col(width = 0.6, alpha = 0.8) +
  
  # Use colors that avoid stereotyping while remaining clear
  scale_fill_manual(values = c("Female" = "#E8B4CB", "Male" = "#87CEEB")) + 
  scale_y_continuous(labels = comma_format()) +
  
  # Add clear data labels to emphasize the scale difference
  geom_text(aes(label = paste0(comma(gender_incident_total))), 
            position = position_dodge(width = 0.9), 
            vjust = 0.5, 
            size = 3, 
            fontface = "bold",
            lineheight = 0.8) +
  labs(title = "Gender and Youth Crime: Different Risks, Different Needs (2024)",
       subtitle = "Understanding why young men and women follow different paths",
       x = "Gender",
       y = "Total Annual Incidents") +
  theme_minimal() +
  theme(plot.title = element_text(size = 12, family = "Arial Black"),
        plot.subtitle = element_text(size = 10),
        legend.position = "none",
        panel.grid.major.x = element_blank(),  
        axis.text = element_text(size = 10))

ggplotly(p5)%>%
  layout(title = list(text = paste0("Gender and Youth Crime: Different Risks, Different Needs (2024)" ,
                                   "<br><sup>Understanding why young men and women follow different paths</sup>")))

```
Slide 7 {data-orientation=columns}
========================================

Row {data-height=200}
-------------------------------------

**Age-by-Age Breakdown: The Detailed Picture of Risk**

**The Developmental Story:** This detailed breakdown reveals how criminal behavior patterns emerge and evolve through the teenage years:

• **Early Teens (10-13):** Lower overall incidents but important warning signs that shouldn't be ignored

• **Peak Risk Period (15-17):** Dramatic increases, especially among young men, reflecting developmental vulnerabilities

• **Late Teens (18-19):** Patterns begin to stabilize but remain elevated, showing the lasting impact of earlier interventions or lack thereof

**Intervention Timing:** This data suggests that the window for most effective intervention may be narrower than previously thought, with ages 13-15 being particularly critical for prevention efforts.

Row {data-height=500}
-------------------------------------

```{r}

# Create a comprehensive view of how risk changes with each year of development
p6 <- ggplot(detailed_age_patterns, aes(x = factor(individual_age), y = total_incidents, fill = gender_category)) +
  geom_col(position = "dodge", alpha = 0.9) +
  
  # Use colors that maintain consistency with previous gender analysis
  scale_fill_manual(values = c("Female" = "#D8A7CA", "Male" = "#7FB3D3")) +
  scale_y_continuous(labels = comma_format()) +
  labs(title = "The Developmental Journey: How Risk Changes Year by Year",
       subtitle = "Each age brings different challenges and opportunities for intervention",
       x = "Age (Years)", 
       y = "Annual Criminal Incidents",
       fill = "Gender") +
  theme_minimal() +
  theme(plot.title = element_text(size = 12, family = "Arial Black"),
        legend.position = "bottom")

ggplotly(p6)%>%
  layout(title = list(text = paste0("The Developmental Journey: How Risk Changes Year by Year" ,
                                   "<br><sup>Understanding why young men and women follow different paths</sup>")))

```
Slide 8 {data-orientation=columns}
========================================

Row {data-height=200}
-------------------------------------

**The Pandemic Paradox: How COVID-19 Created Unexpected Patterns**

**The Pandemic's Contradictory Effects:**

• **2020 - Early Pandemic Surge:** Contrary to expectations, youth crime spiked to decade-high levels as families faced unprecedented stress, economic uncertainty, and social disruption

• **2021 - Lockdown Impact:** This was when restrictions truly took effect, creating the decade's lowest youth crime levels as movement and social interactions were severely limited  

• **Delayed Response:** The data suggests that pandemic stress initially increased youth crime, with the preventive effects of restrictions only appearing later

**Lessons About Social Disruption:** The pandemic taught us that major social changes can have unexpected and delayed effects on youth behavior, with initial stress potentially increasing crime before protective restrictions take effect.

Row {data-height=500}
-------------------------------------

```{r}

# Analyze how the pandemic created two distinct phases of impact
pandemic_period_comparison <- vulnerable_age_groups %>%
  mutate(time_period = case_when(
    Year %in% c(2015:2019) ~ "Pre-Pandemic",
    Year == 2020 ~ "Early Pandemic Surge", 
    Year == 2021 ~ "Lockdown Impact",
    Year %in% c(2022:2024) ~ "Post-Pandemic Recovery"
  )) %>%
  # Remove the average calculation since we want to see the actual 2020 and 2021 data
  filter(!is.na(time_period)) %>%
  group_by(time_period, `Age Group`) %>%
  summarise(period_incidents = if(time_period[1] %in% c("Early Pandemic Surge", "Lockdown Impact")) {
    sum(yearly_incidents)  # For single years, just sum
  } else {
    mean(yearly_incidents)  # For multi-year periods, take average
  }) %>%
  ungroup()

# Visualize the complex pandemic impact
p7 <- ggplot(pandemic_period_comparison, aes(x = factor(time_period, levels = c("Pre-Pandemic", "Early Pandemic Surge", "Lockdown Impact", "Post-Pandemic Recovery")), y = period_incidents, fill = `Age Group`)) +
  geom_col(position = "dodge", alpha = 0.85, width = 0.7) +
  
  # Use a color scheme that reflects the complexity of the pandemic period
  scale_fill_manual(values = c("10-14 years" = "#CCD5AE", 
                              "15-17 years" = "#FFD0D0", 
                              "18-19 years" = "#BED7DC")) +
  scale_y_continuous(labels = comma_format()) +
  labs(title = "COVID-19's Complex and Unexpected Impact on Youth Crime",
       subtitle = "The pandemic created both the decade's peak AND lowest points",
       x = "Time Period", 
       y = "Incidents per Period",
       fill = "Age Group") +
  theme_minimal(base_size = 14) +
  theme(plot.title = element_text(size = 12, family = "Arial Black"),
        axis.text.x = element_text(angle = 45, hjust = 0.5),
        legend.position = "right")
ggplotly(p7)%>%
  layout(title = list(text = paste0("COVID-19's Complex and Unexpected Impact on Youth Crime" ,
                                   "<br><sup>The pandemic created both the decade's peak AND lowest points</sup>")))

```
Slide 9 {data-orientation=columns}
========================================

Row {data-height=200}
-------------------------------------

**Current Concerns: The Troubling 2022-2024 Trend**

**Why These Recent Increases Matter:**

• **Accelerating Growth:** The rate of increase appears to be speeding up, not slowing down

• **Broad-Based Problem:** All age groups are contributing to the increases, suggesting systemic rather than isolated issues

• **15-17 Age Group Driving Growth:** This critical developmental period continues to show the most concerning patterns

• **Time for Action:** The trajectory suggests that without intervention, 2025 could see even higher levels

**The Urgency of Response:** These recent trends indicate that current prevention and intervention strategies may need significant strengthening or reimagining to address the evolving challenges facing Victorian youth.

Row {data-height=500}
-------------------------------------

```{r}

# Focus on the most recent period to understand current challenges
recent_detailed_trends <- vulnerable_age_groups %>%
  filter(Year %in% 2022:2024) %>%
  group_by(Year) %>%
  mutate(annual_total = sum(yearly_incidents),
         age_group_percentage = (yearly_incidents / annual_total) * 100)

# Create a stacked visualization that shows both growth and composition
p8 <- ggplot(recent_detailed_trends, aes(x = factor(Year), y = yearly_incidents, fill = `Age Group`)) +
  geom_col(alpha = 0.9) +
  geom_text(aes(label = format(yearly_incidents, big.mark = ",")), 
            position = position_stack(vjust = 0.5), 
            color = "white", fontface = "bold", size = 3) +
  scale_fill_viridis_d(option = "plasma") +
  scale_y_continuous(labels = comma_format()) +
  labs(title = "Current Challenges: The Concerning 2022-2024 Escalation",
       subtitle = "Understanding why youth crime is increasing and what it means for communities",
       x = "Year", y = "Young People Involved in Criminal Incidents",
       fill = "Age Group") +
  theme_minimal() +
  theme(plot.title = element_text(size = 12, family = "Arial Black"),
        legend.position = "bottom")

ggplotly(p8) %>%
  layout(title = list(text = paste0("Current Challenges: The Concerning 2022-2024 Escalation",
                                   "<br><sup>Understanding why youth crime is increasing and what it means for communities</sup>")))

```
Slide 10
===================================

**Making Sense of It All: What This Data Means for Our Communities**

**The Human Story Behind the Statistics:**

After examining a decade of data representing thousands of young Victorian lives, several clear patterns emerge that should guide our community response:

1. **The Critical Window:** Ages 15-17 represent the highest-risk period, when brain development, social pressures, and limited life experience create perfect conditions for poor decision-making.

2. **Gender-Responsive Needs:** Young men account for approximately three-quarters of incidents, but this doesn't mean young women should be ignored. Each group needs tailored approaches that address their specific risk factors.

3. **Crime Type Focus:** Property crimes dominate youth offending, suggesting that prevention programs should emphasize decision-making skills, understanding consequences, and providing positive alternatives to thrill-seeking behavior.

4. **Recent Escalation:** The post-pandemic increase demands immediate attention, as current trajectories could lead to crisis-level incidents affecting thousands more young lives.

**Moving Forward Together:**

This data tells us that youth crime isn't an inevitable problem—it's a community challenge that responds to targeted, evidence-based interventions. The patterns we see reflect broader social conditions, developmental realities, and the effectiveness (or lack thereof) of current support systems.

**What Communities Can Do:**
- Recognize that every young person involved in crime represents both a failure of prevention and an opportunity for redemption
- Support evidence-based programs that address root causes rather than just responding to symptoms  
- Advocate for sustained investment in youth development, particularly during the critical 15-17 age period
- Create positive opportunities for young people that provide the excitement and belonging they might otherwise seek through risky behavior

**REFERENCES**

Crime Statistics Agency Victoria. (2024). Alleged offender incidents by age and sex, Victoria, 2015–2024 [Data set]. Victorian Government. https://www.crimestatistics.vic.gov.au/crime-statistics/latest-victorian-crime-data/download-data

RStudio. (n.d.). Storyboard layouts. flexdashboard. https://pkgs.rstudio.com/flexdashboard/articles/layouts.html#storyboard

---
